---
title: "FillStrata"
description: >
   Optimal allocation of sample sizes for structure statistics
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{FillStrata}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(struktuR)
```


# Getting started using FillStrata

First we set up some test data with 5 strata

```{r}
s2       <- c(35, 3, 2, 30, 7)         # Variance estimate
N        <- c(100, 100, 500, 50, 150)  # N in each strata
names(N) <- paste("st",1:5,sep="")  # Strata-names
totY     <- c(8800,900,5500,400,300)   # Totals in each strata
g        <- c("grA","grA","grB","grB","grB") # Two groups
```

Next we run the optimal allocation

```{r}
a = FillStrata(s2, N)
```

Tables with strata sample sizes when the total sample is 30, 100 or 300.

```{r}
table(a$filledStrata[1:30])
table(a$filledStrata[1:100])
table(a$filledStrata[1:300])
```

Total variance when total sample size is 300

```{r}
a$sumVar[300]
```


Sample size which ensures a total variance is not larger than 1000

```{r}
table(a$filledStrata[1:(max(which(a$sumVar >= 1000)) + 1)])
```

Total sample size of 200 chosen in advance

```{r}
 table(FillStrata(s2, N, totn = 200)$filledStrata)
```

Total sample size of 250 with a minimum of 40 in each stratum
```{r}
table(FillStrata(s2, N, totn = 250, min_n = 40)$filledStrata)
```

Total sample size of 300 with a minimum of 40 in the first two strata and 50 in the others.

```{r}
table(FillStrata(s2, N, totn = 300, min_n = c(40, 40, 50, 50, 50))$filledStrata)
```

Total sample size of 250 with a maximum of 70 in each stratum 
```{r}
table(FillStrata(s2, N, totn = 250, max_n = 70)$filledStrata)
```

Total sample size of 250 with a maximum of 30 in the first two strata and 100 in the last three

```{r}
table(FillStrata(s2, N, totn = 250, max_n = c(30, 30, 100, 100, 100))$filledStrata)
```





# Using FSmatrix

The following is an example using FSmatrix with a total sample size of 200 chosen in advance. 

```{r}
 m = FSmatrix(s2=s2, N=N, totn=200, group=g)
```
 
There are 3 ways of allocating n in each strata
```{r}
m[,1]$n
m[1,]$Strata
m[[1,1]]
```

Alternativ functions:

```{r}
FillStrata(s2=s2,N=N,totn=200,group=g,returnMatrix=TRUE)[[1,1]]
FillStrata(s2=s2,N=N,totn=200,group=g,returnTable=TRUE)
FStotnAlloc(s2=s2,N=N,totn=200,group=g)
```

Allocation using varLimits

```{r}
vL = c(5000,10000)   # variance limits for groups 
m = FSmatrix(s2=s2,N=N,varLimits=vL,group=g)
m[[1,1]] # Allocated n in each strata
m[3,] # Total variance in strata, group (less than limits) and all.
```

Alternatives:
```{r}
FillStrata(s2=s2,N=N,varLimits=vL,group=g,returnMatrix=TRUE)[[1,1]] 
FillStrata(s2=s2,N=N,varLimits=vL,group=g) 
FSvarAlloc(s2=s2,N=N,varLimits=vL,group=g)
```


Calculate CV from totY in strata from known allocation (20%)

```{r}
m = FSmatrix(s2=s2,N=N,totY=totY,max_n=round(N/5),group=g)
m[5,] # CV in strata, group and all.

```


Alternatives for CV in group:
```{r}
FSmatrix(s2=s2,N=N,totY=totY,min_n=round(N/5),group=g,cvLimits=Inf)[[5,2]]
FillStrata(s2=s2,N=N,totY=totY,min_n=round(N/5),group=g,cvLimits=Inf)
FillStrata(s2=s2,N=N,totY=totY,min_n=round(N/5),group=g,cvLimits=Inf,returnMatrix=TRUE)[[5,2]]
FStotCV(s2=s2,N=N,totY=totY,min_n=round(N/5),group=g)
```


Allocation from CV: 2% and 3% in the two groups

```{r}
m=FSmatrix(s2=s2,N=N,totY=totY,cvLimits=c(2,3),group=g)
m[[1,1]] # Allocated n in each strata
m[5,] # CV in strata, group (less than limits) and all.


```

Alternatives:
```{r}
FillStrata(s2=s2,N=N,totY=totY,cvLimits=c(2,3),group=g,returnMatrix=TRUE)[[1,1]]
FillStrata(s2=s2,N=N,totY=totY,cvLimits=c(2,3),group=g) 
FScvAlloc(s2=s2,N=N,totY=totY,cvLimits=c(2,3),group=g) 
```


Allocation as above, but  totYgroup as intput instead of totY 
```{r}
FSmatrix(s2=s2,N=N,totYgroup=c(10000,5000),cvLimits=c(2,3),group=g)[[1,1]] 
# Alternatives
FillStrata(s2=s2,N=N,totYgroup=c(10000,5000),cvLimits=c(2,3),group=g,returnMatrix=TRUE)[[1,1]]
FillStrata(s2=s2,N=N,totYgroup=c(10000,5000),cvLimits=c(2,3),group=g) 
FScvAlloc2(s2=s2,N=N,totYgroup=c(10000,5000),cvLimits=c(2,3),group=g) 
```

Allocation from CV in groups and overall
```{r}
g2 = cbind(struktuR:::makeDummy(g),1) # Create dummy matrix from g and add extra column
colnames(g2)[3] = "All"
# In Ex4 overall CV is 1.66. Now we set 1.6 as overall limit in addition to "2" and "3" 
m=FSmatrix(s2=s2,N=N,totY=totY,cvLimits=c(2,3,1.6),group=g2)
```


As previous example with totnGroup as new input
```{r}
FSmatrix(s2=s2,N=N,totYgroup=c(10000,5000),cvLimits=c(2,3),group=g,totnGroup=c(10,100))[[1,1]]
# Group 1: Results changed so that total is 10. Group 2 below 100 -> no change
```


As earlier example with min_nGroup as new input


```{r}
FSmatrix(s2=s2,N=N,totYgroup=c(10000,5000),cvLimits=c(2,3),group=g,min_nGroup=c(10,100))[[1,1]]
# Change in group 2 since below 100
```

# Examples use_n option
Create an allocation plan where we select 50%, 25%, 12.5% etc in each stratum
First create a table with the desired sample size choices in each stratum

```{r}
use_n = round(rbind(N, N / 2, N / 4, N / 8, N / 16, N / 32, N / 64, N / 128, N / 256))
#'   use_n
```

Second create the plan using these numbers with a total sample size of 300 
```{r}
table(FillStrata(s2, N, totn = 300, use_n = use_n)$filledStrata)
```

Since this doesn't quite fit we adjust the total sample size up to 313 (stratum 3 is missing 13 to fullfill requirement)

```{r}
table(FillStrata(s2, N, totn = 313, use_n = use_n)$filledStrata)
```

Equivalent with a total sample size of 100
```{r}
 table(FillStrata(s2, N, totn = 100, use_n = use_n)$filledStrata)
```

Strata nr 1 has one too many so we can re-run with a total sample size of 99

```{r}
table(FillStrata(s2, N, totn = 99, use_n = use_n)$filledStrata)
```


Find out prior to running the number that gives the prefect match
```{r}
d <- diff(FillStrata(s2, N, use_n = use_n, corrSumVar = TRUE)$sumVar)
tn <- 1+which(log(d[2:length(d)] / d[1:(length(d)-1)])< -1E-10)
tn
```


Print out all the possibilities  
```{r}
a <- FillStrata(s2, N, use_n = use_n)
for(totn in tn){
  cat("\n----------------------\n      Total n =", totn, "\nTotal variance =", a$sumVar[totn])
  print(table(a$filledStrata[1:totn]))
} 
```

As above but with the optimal alternative (without use_n)

```{r}
aOpt <- FillStrata(s2, N)
for(totn in tn){
    cat("\n\n----------------------\n       Total n =",totn,
         "\nTotal variance =", a$sumVar[totn],
         "\n  Opt variance =", aOpt$sumVar[totn],"\n")
     print(table(a$filledStrata[1:totn]))
     print(table(aOpt$filledStrata[1:totn]))
   }
```

Additional constraint that strata sample sizes are at least 20
Strata sample size sizes between 50% and 100% are not allowed (50% and 100% only)
This is achieved by using both min_n and use_n 

```{r}
table(FillStrata(s2, N, totn = 200, min_n = 20, use_n = use_n[1:2, ])$filledStrata)
```

Ten-fold alternative which invokes printing of "---------#---------#---" during calcs

```{r}
table(FillStrata(s2, 10 * N, totn = 10 * 200, min_n = 10 * 20, use_n = 10 * use_n[1:2, ])$filledStrata)
```

# Examples of use with varLimits and group options

Sample size to ensure the total variance is not larger than 1000 
```{r}

FillStrata(s2, N, varLimits=1000) 
```


Sample size that ensures the total variance for the first two strata and last three strata is not larger than 500 
```{r}
FillStrata(s2, N, varLimits = c(500, 500), group = c(1, 1, 2, 2, 2)) 
```


The total variance in each group given a sample size of 75 in each stratum

```{r}
FillStrata(s2, N, varLimits = Inf, min_n = rep(75, 5), group = c(1, 1, 2, 2, 2)) 
```

Sample size that ensures the total variance for the first group < 100 and the total variance < 1000 

```{r}
md = struktuR:::makeDummy(c(1, 1, 2, 2, 2))
md[, 2] = 1
FillStrata(s2, N, varLimits = c(100, 1000), group = md) 
```

